A general and efficient representation of Ancestral Recombination Graphs

### 2022-09-14 paper.tex

% \section*{Genome ARGs}
There is no fundamental reason, however, that the data structure we use
to describe patterns of genetic inheritance must be in terms
of specific causal events. As we demonstrate in subsequent sections,
there are significant advantages to using
a slightly different formulation, which
we call a ``genome ARG'' (gARG) to distinguish it from the classical
eARG described above.
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance). To recover the local tree for a genomic
position $x$, we traverse the graph rootwards from the leaves
(as in an eARG) and at a particular node, we follow the outbound
edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
Fig.~\ref{fig-arg-in-pedigree}
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
In this diploid setting
(there is no limitation on mating system, ploidy, or age structure
in the gARG encoding),
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty.

% To distinguish from the classical Griffiths eARG,
% we refer to this encoding of an ARG data structure
% as a ``Genome ARG'', or gARG.
% The distinction between nodes representing events
% and genomes may seem pedantic at first, but as we
% see in later sections, we gain a great deal of flexibility
% and computational efficiency by decoupling our representation
% of genetic ancestry from the events that generated it.

@@

% The reader may arrive at the end of this section going, er, so what?
% Try to tackle this head on by getting to the heart of the distinction.
The difference between eARGs and gARGs are subtle, and whether
we choose to interpret nodes as events or genomes
is often of little practical importance.
However, as we show in the remainder of this article,
there are substantial practical gains
from using the gARG encoding, allowing differential
levels of information about the recombination process to be
represented in a computionally efficient way.
The key to this is not in how we interpret nodes, but in how
we represent the process of genetic inheritance.
The critical distinction is that
the gARG encoding allows us to associate
genetic inheritance in non-contiguous chunks of genome
between an ancestor and a descendant.

>>>>

% \section*{Genome ARGs}
There is no fundamental reason, however, that the data structure we use
to describe patterns of genetic inheritance must be in terms
of specific causal events. As we demonstrate in subsequent sections,
there are significant advantages to using
a slightly different formulation, which
we call a ``genome ARG'' (gARG).
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance). To recover the local tree for a genomic
position $x$, we traverse the graph rootwards from the leaves
(as in an eARG) and at a particular node, we follow the outbound
edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
Fig.~\ref{fig-arg-in-pedigree}
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
In this diploid setting
(there is no limitation on mating system, ploidy, or age structure
in the gARG encoding),
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty.

@@

% The reader may arrive at the end of this section going, er, so what?
% Try to tackle this head on by getting to the heart of the distinction.
The difference between eARGs and gARGs are subtle, and whether
we choose to interpret nodes as events or genomes
is often of little practical importance.
However, as we show in the remainder of this article,
there are substantial practical gains
from using the gARG encoding, allowing differential
levels of information about the recombination process to be
represented in a computionally efficient way.
The key to this is not so much in how we interpret nodes
(see Appendix XXX), but specifically in how
we represent the process of genetic inheritance.
The critical distinction is that
the gARG encoding allows us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, something that
cannot be done by storing recombination breakpoints.
